Gradual Soundness: Lessons from Static Python

### 2022-04-26 pj-7.1/response.txt

TODO
- write a quick thank-you and summary of changes
- add section + page numbers to the responses

> Review #45A
> ===========================================================================
>
> Paper summary (for authors)
> ---------------------------
> The paper makes a contribution to gradual typing in theory and practice. The
> subject language is Static Python, which adds a compile-time type discipline
> to Python.
>
> On the theory side, the paper describes a formal model created with an
> explorative tool for semantics modeling and prototyping. This model includes a
> static semantics and a dynamic semantics. The dynamic semantics is tested
> against a regression test suite.
>
> On the practical side, the paper describes the design of Static Python as
> well as experiments with an implementation of Static Python. This
> implementation allegedly improves the throughput of a closed source
> implementation of a web server by 3.6%. It improves the run time of several
> microbenchmarks significantly.
>
> Assessment and comments (for authors)
> -------------------------------------
> ## strengths
>
> * The main strength of the paper is clearly the favorable evaluation of a
>   particular datapoint in the design space of gradual typing.
> * The formal model enabled the discovery of problems in a industrial strength
>   implementation of Static Python, at least in its type checker.
>
> ## weaknesses
>
> * the most significant evaluation is on a closed source program
> * the microbenchmarks potentially gain more efficiency from progressive
>   primitive types as most of them are computationally intensive.
> * the theoretical evaluation is only probabilistic with unclear probability
>
> ## comments
>
> I like the paper, it is generally well readable. There are a few places,
> where I'd expect more guidance from the authors, rather than expecting the
> reader to interpolate.
>
> ### Table 1
>
> The caption is misleading. The data is the cost of creating and using
> shallow/concrete dictionary types in typed and untyped contexts.  I was
> initially confused by the 0 entries. I'd prefer you put --- to indicate that
> no casts is needed.  You should indicate that $N$ is the length of the
> dictionary.

@@

It is always a fixed-size pointer.

> ### p10, 2
>
> *only if*
> Is that an `iff`?

No, the "only if" is intentional. An "iff" would be correct, but the paragraph
in which this sentence appears is about one direction in particular.

> ### p10, -17
>
> *Redex's random testing ... to check type soundness*
>
> That seems a bit shaky. What's the baseline / ground truth / oracle for these
> tests?
> In the end, all that testing only gives you a probabilistic guarantee.
> Can you give an estimate of this probability? (I mean, quantitative)

@@

> ### p12, fig 3, 4
>
> * `str` should be italic
> * why is `object` part of the expression language?

Fixed `str` and removed `object`.

(`object` was there as a simple instance of the `Object` type. Since it's not
needed to explain type boundary issues, we can remove it.)

> Line 6: `C.f(expr)`
> * looks like a static method call to me, is that right? That's usually
>   trivial compared with the dynamic dispatch of a real method call.
> * how do you invoke a method on an object?

@@

Good point, we've switched `S` and `T`.

> ### table 2
>
> Misleading caption: this is about the cast from Any to an evaluation type.

We have reworded the caption.

> ### p14, fig 5
>
> * rule `M-App` seems too limiting. Literally read, it's not possible to
>   invoke inherited methods because the method `f0` must be defined in class
>   `C0` in which it is defined. This rule affirms the class-method conjecture
>   from above.

@@

We had envisioned the lookup `Env(C0)` as searching both `C0` and its parents
for all visible fields and methods.

The revised figure writes `Env*(C0)` instead and includes a note to this effect.
(The revision also uses `\Gamma` instead of `Env`)

> * the `object` rule for subtyping contradicts Table 2, which says that
>   *objects accepts any value expect primitives*. So `S0` must not be a
>   primitive; but primitives are not part of the model. This is not
>   consistent!

@@

Method calls in Redex are standard; they work like calls in (Static) Python.
The confusion here is due to our over-simplified model in the submission, which
gave the impression that only a class can receive a method call.

> ### p24
>
> references [23] and [24] and [41] have broken DOIs

Fixed.

> * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
>
>
> Review #45B
> ===========================================================================
>
> Paper summary (for authors)
> ---------------------------
> The contribution of the paper is the following:
>
> - The paper describes Static Python, a project developed at Instagram to add
>   types to Python code in a gradual manner, with the main objective of
>   improving performance. The project has already shown its effectiveness. The
>   authors (informally) describe both the language (the gradual type system)
>   and how this is implemented to achieve better performance.
>
> - The authors have developed an implementation in PLT redex of a subset of
>   Static Python, focusing on features related to gradual typing. Using this
>   implementation, they did an extended testing to show:
>   1) that this implementation is sound
>   2) that this implementation gives the same results of Static Python.
>
> Assessment and comments (for authors)
> -------------------------------------
> Assessment: The contributions of the paper summarized above are certainly
> useful for the PL community, so I support acceptance. The empirical evaluation
> seems to me (but I am definitely not an expert in such kind of contributions)
> well-driven and adequate. I have some reservations on the formalization,
> explained in the comments below, which should be solved/improved in the final
> version.
>
> Detailed comments:
>
> Page 3 Mechanization: you should explain better what do you mean by
> "soundness". I thought of standard soundness in the sense that "well-typed
> programs do not go stuck", but here you say "property-based soundness tests",
> what do you mean more precisely?

@@

Section 4 in the revised paper gives more details.

In short, we started with the property "well-typed programs do not get stuck"
and tested it on thousands of terms generated by Redex.

> Section 4.1 line 4: In the formal syntax in Figure 3, Object seems not to be
> an instance of C, so class C(C) seems not to include the case C = Object.
> In the "Other expressions" you mention object fields reads and writes, but in
> the formal syntax in Figure 3 you have C.x and C.x = expr, I was expecting
> expr.x and expr.x = expr' instead.

@@

We've made several changes in response:

- `Object` is no longer part of the grammar
- The grammar comes with a note that the set `C` must contain `Object`
- Field reads and writes use `expr` instead of `C` (example: C.x is now expr.x)

> Figure 3: it would be better to use a more abstract style with metavariables
> as it is customary now, e.g.
> prog = stmt_1, ..., stmt_n
> expr = ... [x_1:expr_1, ..., x_n:expr_n] | expr[expr'] |

@@

(Note that `x` and `f` are undifferentiated variable names. The model uses `x`
 in some definitions and `f` in others as hints to readers, but it's fine to
 name a function `x_2` and so on.)

> Also, Object seems to be the top type, hence also Bool and Int are seen as
> classes, right?

Yes, Bool and Int are technically classes. The paper says little about this,
but the Redex model (and Static Python) give them full support as classes.

> In rules D-Set and CD-Set, the type None seems to be used as a "Unit" type.
> The fact that these two types are identified seems strange.

### 2022-04-26 pj-7.1/paper.tex
edit cleanup

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
assertions, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is available at:
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth---using our
own expertise---to look for non-conformance.  Second, 
we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
type soundness.
Specifically, we asked Redex to generate well-formed terms (using its default
algorithm) and checked whether the well-typed terms that terminated within a
fixed time budget (1,200 expressions and 11,000 programs in a typical run)
produced only well-typed values and allowed errors.
% \footnote{Property-based testing helped us find and fix several issues in the
% model.  None of issues had implications for \SP{}'s soundness.}
Finally, we employed the well-established method of testing end-to-end
conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
We translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,
which suggests that the model conforms to actual \SP{}.

@@

By contrast to our Redex model, the typing judgment in~\cref{f:eval-type} is
not syntax directed.
The rules do not describe a typechecking algorithm.
Instead these rules declare various conclusions and their preconditions to
illustrate how \SP{} reasons about types---dynamic or otherwise.

@@

Almost all ratios are less than one, showing that \SP{} achieves a speedup over
Python (typed avg. $0.8$, refined avg. $0.3$).
The exceptions are the unrefined versions of \bmname{fannkuch} and
\bmname{nbody}, which is because these benchmarks focus
on number-crunching and \SP{} optimizes arithmetic only for primitive types.
Changing a few types in \bmname{fannkuch} to primitives
gives an immediate speedup (ratio = $0.90$).

@@

Concrete types originated in Thorn~\cite{bfnorsvw-oopsla-2009,wnlov-popl-2010}
and StrongScript~\cite{rzv-ecoop-2015}, two early languages that support typed
and untyped objects.
Thorn is a new scripting language for the JVM; StrongScript is designed to
interoperate with JavaScript.
Concrete types came to \SP{} by way of Nom~\cite{mt-oopsla-2017} and its
successor MonNom~\cite{mt-oopsla-2021}.
The impressive performance figures in the original Nom paper inspired the \SP{}
team to adapt the method.
But whereas MonNom has been exploring designs for generic concrete types that
support (a flexible form of) the gradual guarantees~\cite{svcb-snapl-2015},
\SP{} opts for a simpler and more rigid design~(\cref{s:type-dynamic}).

>>>>

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
assertions, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is available at:
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth---using our
own expertise---to look for non-conformance.  Second, 
we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
type soundness.
Specifically, we used Redex to generate well-formed terms (using its default
algorithm) and checked whether the well-typed terms that terminated within a
fixed time budget (1,200 expressions and 11,000 programs in a typical run)
produced only well-typed values and allowed errors.
% \footnote{Property-based testing helped us find and fix several issues in the
% model.  None of issues had implications for \SP{}'s soundness.}
Finally, we employed the well-established method of testing end-to-end
conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
We translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,
which suggests that the model conforms to actual \SP{}.

@@

By contrast to our Redex model, the typing judgment in~\cref{f:eval-types} is
not syntax directed.
The rules do not describe a typechecking algorithm.
Instead these rules declare various conclusions and their preconditions to
illustrate how \SP{} reasons about types---dynamic or otherwise.

@@

Almost all ratios are less than one, showing that \SP{} achieves a speedup over
Python (typed avg.~$0.8$, refined avg.~$0.3$).
The exceptions are the unrefined versions of \bmname{fannkuch} and
\bmname{nbody}, which is because these benchmarks focus
on number-crunching and \SP{} optimizes arithmetic only for primitive types.
Changing a few types in \bmname{fannkuch} to primitives
gives an immediate speedup (ratio = $0.90$).

@@

Concrete types originated in Thorn~\cite{bfnorsvw-oopsla-2009,wnlov-popl-2010}
and StrongScript~\cite{rzv-ecoop-2015}, two early languages that support typed
and untyped objects.
Thorn is a scripting language for the JVM; StrongScript is designed to
interoperate with JavaScript.
Concrete types came to \SP{} by way of Nom~\cite{mt-oopsla-2017} and its
successor MonNom~\cite{mt-oopsla-2021}.
The impressive performance figures in the original Nom paper inspired the \SP{}
team to adapt the method.
But whereas MonNom has been exploring designs for generic concrete types that
support (a flexible form of) the gradual guarantees~\cite{svcb-snapl-2015},
\SP{} opts for a simpler and more rigid design~(\cref{s:type-dynamic}).
